Theoretical prediction of disrupted Min oscillation in flattened \emph{Escherichia coli}

### 2015-03-27 paper/plos.tex
a few edits with Jeff alongside (todo is done!) Tested everything on bennet.

\red{Huh}

\fixme{
  First referee:
  \begin{enumerate}
  \item Done%% \blue{Done} Add discussion of ``platform'' --- wrote our own code,
    %% publically available (cite github?).
  \item Done
    %% \blue{Done} Make 3D geometries available? Cite that the publically
    %% available code can generate them?
  \item Done%%Claims differential equations aren't needed.
  \item Doen%%Claims wild-type movie-ish plot not needed.
  \item Points out problems with Huang 2003:
    \begin{enumerate}
    \item \blue{David.  I've added some sentences discussing the
      starry night effect and membrane diffusion blurring out the
      stars.  I have some confusion however about the diffusion
      lengths.}  Lacks diffusion on the membrane. Supported by Meacci
      2006, which experimentally tries to measure diffusion on
      membrane, has a tough time, but shows that definitely diffuses.
    \item \blue{David} Underestimates diffusion in cytoplasm by a factor of 5-6.5.
      Supported by Meacci 2006, who concludes a rate of 16 microns
      squared per sec for MinD and 10 microns squared per sec for
      MinE.
    \item \blue{David} Uses MinD-ADP -> MinD-ATP that is 3 times lower than
      experimental bound. Halatek uses ADP->ATP of 1/6 s-1. They cite
      Meacci 2006, who state that there is a lower bound on reference
      time of 3 seconds, so this constant has an upper bound of 1/3
      s-1.
    \item \blue{David} (Still debated) No evidence that DE recruits to membrane.
    \item Done%% Works in only a limited part of parameters phase space, and
      %% could not account for important things like
      %% temperature-independence.
    \end{enumerate}
  \item Done%% \blue{Done} Height should be bigger to compare with experiment.  0.4-0.6
    %% microns.  The 0.25 micron height of the cells used in this study
    %% does not represent the experimental settings well. The simulations
    %% are based on the flattened cell shapes reported from the
    %% experiments by Mannik et al 2009 and 2012. In these experimental
    %% studies, E. coli were shown to be able to grow in channels as low
    %% as 0.6 micron in hard silicon structures, and no more than 0.4
    %% micron (see Mannik 2009). Indeed, 0.25 micron channels were
    %% constructed with PDMS ceilings, which is a rather flexible polymer
    %% allowing bacteria to deform it to effectively higher
    %% channels. Thus the simulations would be more realistic if
    %% performed in the range of 0.4-0.6 micron thickness.

    %% \emph{If you can't make it grow in a .3 micron (rigid) silicon channel,
    %%   then we and you observe it growing in a .25 micron flexible
    %%   channel, it presumably must be able to grow because it deformed
    %%   the channel to larger than .3 microns.}
  \item Done%%\blue{David} Starry night in the wild-type cells is a problem.
  \item Done%%Better explanation of correlation function (which is
    %%interesting!).
  \item Done%%Shape A is the interesting bit in our main figure.
  \item Done%%Would like more analysis of stochastic locations in different
    %%geometries.  Clusters?  Could add more geometries?
  \item Done%%Bold conclusions about flattening.  Talk about both width and
    %% thickness.  Compare Halatek 2012.  Wants systematically more
    %% geometries again.
  \item Done%%Is doubtful of our correlation plots relative to those of
    %%Fange and Elf 2006.
  \item Done\blue{David. Read the refferee report with regard to this, and
    not sure what he wants...} Clarify (esp in conclusions) what we
    mean by flattening and irregular shape.  ``breaking the
    cylindrical symmetry and flattening the cells.'' versus ``breaking
    the mirror plane symmetry of the cross-section of a flattned
    cell.''
  \item Done%%[TODO suggestion] Switch to Halatek and Frey
  \item Done%%[TODO suggestion] Thoroughly explor shape parameter space
  \item Done%%[TODO suggestion] Check robustness of our conclusions with
    %%respect to changes in (reaction-diffusion?) parameters.
  \item Done%%Check order of figures and their references.
  \item Done%%%% \blue{David.  I cited the Corbin paper (although he's got a
    %% typo because Corbin 2008 is after the fange 2006 paper. I could
    %% cite other experimental papers, but the sentence in question
    %% specifically refers to changes made to the Huang model, and the
    %% papers cited are the main reasons that those changes have been
    %% made.)} Cite biological (non-computational) works, that the
    %% e.g. Corbin 2008.
  \end{enumerate}
  Second referee:
  \begin{enumerate}
  \item Done%%Nitpicks about language of deterministic model versus
    stochastic model.  (Read carefully)
  \item Done%%\blue{David} Explain in greater detail the obviousness (maybe) that the
    %% deterministic approach doesn't have fluctuations.  RESPOND:
    %% chaotic systems do deterministically result in fluctuations.
    %% Explain this somewhere, and discuss triviality of the conclusion
    %% that deterministic is wrong because it does not exhibit these
    %% irregularities.
  \item Would like temporal averages over cell-cycle time scales.
  \item More discussion of how flattening leads to spatial
    irregularities.  Compare with large nearly spherical e. coli Shih
    2005.  See also Halatek and Frey 2012 for rotating oscillation
    axis due to destabilization of short modes.  How does this connect
    with our stability analysis?
  \item Done%%\blue{David} Should mention more recent revisions of the model.
  \item Done%%\blue{David} Fange and Huang do not use the same reaction-diffusion
    %%equations.
  \item Done%%%% \blue{David.  I read the referee's note about this.  Not sure
    %% what he/she wants.}  Clarify statement about 1000 proteins and
    %% stochastic fluctuation.  Reword about a continuum description
    %% omitting information?
  \item Done%%%% \blue{Done} Do not call Huang parameters ``wild type values.''  They are
    %% fitted parameters.
  \item Done%%\blue{Done?  I think our additions to the Methods section
    %% addressed this?}  Add methods section for numerics
    %% (deterministic).
  \item Done%%\blue{Done} Add details of the linear stability analysis.
  \item Done%%\blue{David} Still need to fix periods on correlation plots.
    %% \blue{About this point - I updated the data and it seems the
    %%   result is defferent from before - the stadium shapes seem to be
    %%   temporally correlated for much longer compared with the shape A
    %%   than before.}
  \end{enumerate}
}

@@

\blue{5a,b,c,d}Several models
\cite{fange2006noise,bonny2013membrane,halatek2012highly} modify
Huang's original model to more accurately reflect experimental
findings, which were published after his 2003
paper~\cite{meacci2006mobility,loose2011min}. These models introduce
diffusion along the membrane that is two to three orders of magnitude
slower than that within the cytoplasm, and increase the rate of
cytoplasmic diffusion by a factor of about five.  They also increase
or assume instantaneous~\cite{bonny2013membrane} the nucleotide
exchange rate.  In addition, they revise the process of MinD
recruitment to the membrane, prohibiting MinD from attaching to
MinD-MinE complexes there, behavior that is experimentally unresolved.
The model of Bonny \emph{et al.} allows MinE to remain independently
attached to the membrane after dissasociating with
MinD~\cite{bonny2013membrane}.
%

There are only on the order of 1000 MinD proteins in a given cell,
which is significantly less than the number of gridpoints used in our
simulations, suggesting that a continuum description will break down
and stochastic variation plays a significant role.  To treat this,
there have been a number of variations of the Huang 2003 model that
stochastically simulate the similar reaction-diffusion
equations~\cite{fange2006noise, kerr2006division}.  These studies
largely confirm the results of Huang's deterministic model when
applied to the wild-type, pill-shaped phenotype.  However, the
stochastic models are slightly more successful in predicting
experimentally observed oscillations in round cell
phenotypes~\cite{fange2006noise, huang2004min}, and they enable
prediction of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

@@

\blue{(5a)} Both Fig.~\ref{image-p} and its associated movies (see
supplementary material), show that the stochastic model exhibits a
``starry night'' effect characterized by spatially fixed points of
protein density build up, as recruitment leads to clusters of MinD
forming on the wall and then subsequently dissipating.  This results
from the lack of inclusion of diffusion along the membrane in Huang's
model, which would allow these recruitment clusters to spread out.
Experiments have confirmed that diffusion along the membrane does
occur, albeit with a rate two orders of magnitude lower than that in
the cytoplasm~\cite{meacci2005min}.  The `stars' in the effect
typically last for around 10 seconds. The membrane-bound proteins will
travel a distance of around one micron during one of these periods of
time, enough to spread out across the polar section of the one micron
diameter wild-type pill cells.  As we see below, the flattened,
irregular cells are large enough so that the one micron blur would not
completely eliminate the `starry night' variation in maxima position
effect.

@@

\subsection*{Comparison of Experimental and Stadium Shapes}
%\label{sec:model-method-shapes}
As described in Section~\ref{sec:model-method-shapes}, \fixme{this
  link?} we will focus on simulation of four illustrative flattened
cell shapes: two shapes created to replicate those shapes observed
experimentally by Mannik \emph{et al.}~\cite{mannik2012robustness}
(shape~\emph{A} and shape~\emph{B}), and for comparison two `stadium'
shapes that have the identical aspect ratio, thickness and volume
(stadium \emph{A} and stadium \emph{B}), shown in
Figure~\ref{randst-plot-ave}.  All of the flattened cells have lost
the rotational symmetry of the orginal pill shapes, but while
shape~\emph{A} and shape~\emph{B} have only one mirror plane symmetry
(that is normal to the flattened plane), stadium~\emph{A} and
stadium~\emph{B} have three.  This allows us to distinguish between
the effect of flattening the cell and the cell shape's irregularity
and asymmetry.  For each of these shapes (in addition to the wild-type
pill shape discussed previously) we have simulated in excess of 15000
seconds (over 4 hours) of evolution of the MinD system using the
deterministic version of Huang's model~\cite{huang2003dynamic}, and in
excess of 100000 seconds (over 27 hours) using the stochastic model.

@@

\section*{Conclusion}
We find agreement between our stochastic simulations and experiments
showing disrupted bipolar MinD oscillation in asymmetric flattened
E. coli cells~\cite{mannik2012robustness}.  As observed
experimentally, our simulations predict that MinD maxima will form in
a spatially irregular sequence of locations.  This result builds upon
existing results showing that this model and its variants are
effective in a variety of wild-type and mutated cell
shapes~\cite{fange2006noise, varma2008min, kruse2007experimentalist},
and reinforces that stochastic variations of Huang's 2003
model~\cite{fange2006noise, kerr2006division} has considerable
predictive power beyond the standard wild-type cell.  \fixme{Discuss
  which of these are likely to map to more recent models.}

%
%% We have simulated flattened cell shapes similar to those observed
%% experimentally by Mannik \emph{et al.}, with a stochastic and a
%% deterministic version of Huang's widely used reaction-diffusion model.
%% %
%% We find that the stochastic model predicts experimentally accurate
%% qualitative behavoir in both the wild-type, pill-shaped cells (as
%% found in previous studies~\cite{fange2006noise, varma2008min,
%%   kruse2007experimentalist}), as well as in the new, flattened,
%% irregularly shaped cells.  These flattened cells invariably show
%% spatially irregular formation of maxima as has been observed
%% experimentally.

In contrast to the stochastic model, the original deterministic
version of the model~\cite{huang2003dynamic} predicts behavior that
contradicts experimental observations.  Specifically, the
deterministic model predicts robust and periodic bipolar oscillation
in irregularly shaped cells.  Thus we conclude that this method is
inconsistent with experiment, and cannot be relied upon for
predictions of the behavior of the MinD system.  This is the most
clear example of the inadequacy of the deterministic simulation method
in reproducing experimental behavior of the MinD system to date.
Until now the results of deterministic and stochastic simulation have
largely coincided, with the stochastic method showing minor
differences in behavior when compared to deterministic
models~\cite{kerr2006division, fange2006noise, huang2004min,
  kruse2007experimentalist}.  The regular oscillations in the
deterministic case also allow us to conclude that the model is
non-chaotic.  Because of their similar construction, we suspect that
recent variants of the
model~\cite{bonny2013membrane,fange2006noise,halatek2012highly} would
also prove to be non-chaotic.

>>>>

@@

Several models \cite{fange2006noise, bonny2013membrane,
  halatek2012highly} modify Huang's original model to more accurately
reflect experimental findings, which were published after his 2003
paper~\cite{meacci2006mobility, loose2011min}. These models introduce
diffusion along the membrane that is two to three orders of magnitude
slower than that within the cytoplasm, and increase the rate of
cytoplasmic diffusion by a factor of about five, reflecting an
experimental measurement of these
parameters~\cite{meacci2006mobility}.  They also increase or assume
instantaneous~\cite{bonny2013membrane} the nucleotide exchange rate.
In addition, they revise the process of MinD recruitment to the
membrane, prohibiting MinD from being recruited to the membrane by
MinD-MinE complexes, behavior that has not been observed
experimentally.  The model of Bonny \emph{et al.} allows MinE to
remain independently attached to the membrane after dissasociating
with MinD~\cite{bonny2013membrane}.
%

There are only on the order of a thousand MinD proteins in a given
cell, which is fewer than the number of gridpoints used in our
simulations, suggesting that the deterministic continuum
description--which requires fractional numbers of proteins at each
grid point---may break down, and that stochastic behavior may play a
significant role.  To treat this, there have been a number of
variations of the Huang 2003 model that stochastically simulate the
similar reaction-diffusion equations~\cite{fange2006noise,
  kerr2006division}.  These studies largely confirm the results of
Huang's deterministic model when applied to the wild-type, pill-shaped
phenotype.  However, the stochastic models are slightly more
successful in predicting experimentally observed oscillations in round
cell phenotypes~\cite{fange2006noise, huang2004min}, and they enable
prediction of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

@@

Both Fig.~\ref{image-p} and its associated movies (see supplementary
material), show that the stochastic model exhibits a ``starry night''
effect characterized by spatially fixed points of protein density
build up, as recruitment leads to clusters of MinD forming on the wall
and then subsequently dissipating.  This effect is strengthened by the
lack of inclusion of diffusion along the membrane in Huang's model,
since this diffusion could allow these recruitment clusters to spread
out.  Experiments have confirmed that diffusion along the membrane
does occur, albeit with a rate two orders of magnitude lower than that
in the cytoplasm~\cite{meacci2005min}.  The `stars' in the effect
typically last for around 10 seconds. The membrane-bound proteins will
diffuse by a distance of around one micron during this time, which is
enough to spread out across the polar section of a one micron diameter
wild-type pill cell.  As we see below, the flattened, irregular cells
are larger so that the one micron blur would not completely eliminate
this `starry night' effect.

@@

\subsection*{Comparison of Experimental and Stadium Shapes}
%\label{sec:model-method-shapes}
As explained above, we focus on simulation of four illustrative
flattened cell shapes: two shapes created to replicate those shapes
observed experimentally by Mannik \emph{et
  al.}~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two `stadium' shapes that have the
identical aspect ratio, thickness and volume (stadium \emph{A} and
stadium \emph{B}), shown in Figure~\ref{randst-plot-ave}.  All of the
flattened cells have lost the rotational symmetry of the orginal pill
shapes, but while shape~\emph{A} and shape~\emph{B} have only one
mirror plane symmetry (that is normal to the flattened plane),
stadium~\emph{A} and stadium~\emph{B} have three.  This allows us to
distinguish between the effect of flattening the cell and the cell
shape's irregularity and asymmetry.  For each of these shapes (in
addition to the wild-type pill shape discussed previously) we have
simulated in excess of 15000 seconds (over 4 hours) of evolution of
the MinD system using the deterministic version of Huang's
model~\cite{huang2003dynamic}, and in excess of 100000 seconds (over
27 hours) using the stochastic model.

@@

\section*{Conclusion}
We find agreement between our stochastic simulations and experiments
showing disrupted bipolar MinD oscillation in asymmetric flattened
E. coli cells~\cite{mannik2012robustness}.  As observed
experimentally, our simulations predict that MinD maxima will form in
a spatially irregular sequence of locations.  This result builds upon
existing results showing that this model and its variants are
effective in a variety of wild-type and mutated cell
shapes~\cite{fange2006noise, varma2008min, kruse2007experimentalist},
and reinforces that stochastic variations of Huang's 2003
model~\cite{fange2006noise, kerr2006division} has considerable
predictive power beyond the standard wild-type cell.

In contrast to the stochastic model, the original deterministic
version of the model~\cite{huang2003dynamic} predicts behavior that
contradicts experimental observations.  Specifically, the
deterministic model predicts robust and periodic bipolar oscillation
in irregularly shaped cells.  Thus we conclude that this method is
inconsistent with experiment, and cannot be relied upon for
predictions of the behavior of the MinD system.  This is the most
clear example of the inadequacy of the deterministic simulation method
in reproducing experimental behavior of the MinD system to date.
Until now the results of deterministic and stochastic simulation have
largely coincided, with the stochastic method showing minor
differences in behavior when compared to deterministic
models~\cite{kerr2006division, fange2006noise, huang2004min,
  kruse2007experimentalist}.  The regular oscillations in the
deterministic case also allow us to conclude that the model is
non-chaotic.  Because of their similar construction, we anticipate
that recent variants of the model~\cite{bonny2013membrane,
  fange2006noise, halatek2012highly} would also prove to be
non-chaotic.
